Capturing the geometric structure of our experiences and how we remember them

### 2018-07-24 main.tex
memorability by event and topics

To visualize the relationship between the video and recall event models, we embedded the video and average recall models into a 2D space (using the UMAP dimensionality reduction algorithm, for details see \ref{sec:methods}) where the points represent video/recall events and the distance between them represents their similarity in ``topic space'' (Fig.~\ref{fig:trajectory}a).   We observed that the two models have a very similar temporal evolution and geometric structure. To further quantify this correspondence and to characterize how participants navigated through the space during recall, we analyzed the angle between successive recall events. We created a grid of evenly spaced points (.25 units) in the 2D ``topic space''. For each point on the grid, we drew a circle (radius=.25) around the point and grouped together all recall transitions that intersected the circle (across subjects). To visualize the average angle, we converted each transition angle to a unit vector and then averaged the vectors together. To assess consistency in the direction of the recall transition across participants, we performed a Rayleigh test (p<.001, corrected at p<.05 using permutation procedure described in \ref{sec:methods}). We found that on average, participants' recalls flowed along the same path as the video model (Fig.~\ref{fig:trajectory}b). Each participant's recall path can be seen in Fig.~\ref{fig:trajectory}c.

@@

\begin{figure}[t!]
\centering
\includegraphics[width=1\textwidth]{figs/5_topics.pdf}
\caption{\small \textbf{Recall shape analysis and video/recall content.} A 2-dimensional embedding of the best fitting line segment model.  The points outlined in black represent and connected by black line segments represent the critical recall events (or ``hinge-points'') of the best fitting model colored according to their matching video event.  The points not outlined in black are individual participant recall event topic vectors colored by the most similar video event. The kernel density plot represents the distribution across all points.  The word clouds for hinge-point (left: video; right: recall) were created by plotting the top 100 words from topic model that were ``activated'' during a particular event.}
\label{fig:topics}
\end{figure}

>>>>

To visualize the relationship between the video and recall event models, we embedded them into a 2D space (using the UMAP dimensionality reduction algorithm, for details see \ref{sec:methods}) where the points represent video/recall events and the distance between them represents their similarity in ``topic space'' (Fig.~\ref{fig:trajectory}a).   We observed that the two models have a very similar temporal evolution and geometric structure. To further quantify this correspondence and to characterize how participants navigated through the space during recall, we analyzed the angle between successive recall events. We created a grid of evenly spaced points (.25 units) in the 2D ``topic space''. For each point on the grid, we drew a circle (radius=.25) around the point and grouped together all recall transitions that intersected the circle (across subjects). To visualize the average angle, we converted each transition angle to a unit vector and then averaged the vectors together. To assess consistency in the direction of the recall transition across participants, we performed a Rayleigh test (p<.001, corrected at p<.05 using permutation procedure described in \ref{sec:methods}). We found that on average, participants' recalls flowed along the same path as the video model (Fig.~\ref{fig:trajectory}b). Each participant's recall path can be seen in Fig.~\ref{fig:trajectory}c.

@@

\subsubsection{Memorability by event and topics}
Another advantage of representing naturalistic stimuli/memory using this approach is that the event models can be mapped back to the language used to fit the model(s).  In this next analysis, we 1) quantified the memorability of each video event and 2) used memorability to plot the top words for the most and least memorable scenes, and across the entire video.

To measure video event memorability, we computed the correlational distance between each video event and the closest recall event and then averaged those values within participant. We used this measure of memory to create a weighted average over all video events, where video events with better memory were weighted more heavily. Then, we extracted the top 200 words and created ``wordles'' from the most memorable scenes of the video (see Methods for details). The result reveal that scenes containing "Sherlock" and "John" are highly memorable, and also "medium", "street" and "baker" (Fig.~\ref{fig:topics}b). To find words related to the least memorable scenes, we inverted the "memorability" weights and extracted the top 200 words.  Scenes containing "Mike" and "Molly" were least remembered, and also words like "conference", "hospital" and "psychotherapist" (Fig.~\ref{fig:topics}b).

Then, we took a closer look at the themes contained in the top/bottom 3 events. We used the "memorability" weights described above to select the top/bottom events. We performed the same analysis as described above, but separately for the video event and average recall event vectors representing each of the top/bottom 3 remembered events. In Figure~\ref{fig:topics}c, the trajectory represents a low-dimensional embedding of the video model and the large colored dots represent the video events. Each of the smaller colored dots represents the distribution of individual recall events over participants, colored by the closest video event.

\begin{figure}[t!h]
\centering
\includegraphics[width=1\textwidth]{figs/5_topics.pdf}
\caption{\small \textbf{Topic vector memorability.} A 2-dimensional embedding of the best fitting line segment model.  The points outlined in black represent and connected by black line segments represent the critical recall events (or ``hinge-points'') of the best fitting model colored according to their matching video event.  The points not outlined in black are individual participant recall event topic vectors colored by the most similar video event. The kernel density plot represents the distribution across all points.  The word clouds for hinge-point (left: video; right: recall) were created by plotting the top 100 words from topic model that were ``activated'' during a particular event.}
\label{fig:topics}
\end{figure}
